---
title: "Air Quality Classification"
author: "Rithik Ganesan"
date: "2025-06-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(tidyverse)
library(ggplot2)
library(class)
library(caret)
```

## Problem Definition
Air quality is becoming an increasingly prevalent issue around the world. In this project I will attempt to create a classification model that correctly classifies a city as industrial/residential using readings of air quality in the area.

I obtained this data from Youssef Elebiary on Kaggle. However, this data set is published under the MIT License, allowing free use for both personal and commercial purposes. Attribution is encouraged but not required.

Link to Dataset: https://www.kaggle.com/datasets/youssefelebiary/industrial-residential-air-quality-classification/data

## Data Wrangling
First and foremost, let's load in and take a look at the format of the data.
```{r load}
airDF <- read_csv("Pollution.csv",show_col_types = FALSE)
summary(airDF)
colSums(is.na(airDF))

```
After inspecting the properties of the data and ensuring data integrity, we can visualize the data to familiarize ourselves with the variables and their impact. 

```{r visual}

avg_by_city <- airDF %>%
  group_by(City) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

# Reshape to long format for faceted plotting
avg_long <- pivot_longer(avg_by_city, cols = -City, names_to = "Variable", values_to = "Average")

# Plot with one facet per variable
ggplot(avg_long, aes(x = City, y = Average, fill = City)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(title = "Average Value by City for Each Variable", x = "City", y = "Average Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```
These visuals make it easy to spot similarities and differences in air quality levels in industrial cities compared to residential cities. One observation worth highlighting is that the industrial cities seem to all have high CO, NO2, PM10, PM2.5, and SO2. On the other hand, residential areas all have relatively lower levels.

Now we must wrangle our data so that it is easy to integrate into multiple classifaction models.
```{r wrangle}
airDF$Date <- NULL
airDF$City <- NULL
airDF$Type <- ifelse(airDF$Type == "Industrial", 1, 0)
head(airDF,5)
```
We remove the date column as date will not play a part in our classification. The city column is also removed to ensure that the model will not be "cheating" by having access to the city it is trying to classify. We then ensure our outcome variable is numerical and preview the head of the data to verify our work. 
We then split the data into training and testing splits as well as seperating factors and outcomes.
```{r pressure, echo=FALSE}
split <- sample(seq_len(nrow(airDF)), size = 0.8 * nrow(airDF))
trainDF <- airDF[split, ]
testDF  <- airDF[-split, ]

train_predictors <- trainDF[ , names(trainDF) != "Type"]
test_predictors  <- testDF[ , names(testDF) != "Type"]

train_outcome <- trainDF$Type
test_outcome  <- testDF$Type

```

## K-Nearest Neighbors
We will first create a KNN classification model and analyze the results.
```{r knn}
set.seed(77)
train_scaled <- scale(train_predictors)
test_scaled  <- scale(test_predictors, center = attr(train_scaled, "scaled:center"), 
                      scale = attr(train_scaled, "scaled:scale"))
k_values <- 1:20
accuracies <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  pred_i <- knn(train = train_scaled,
                test = test_scaled,
                cl = train_outcome,
                k = k_values[i])
  accuracies[i] <- mean(pred_i == test_outcome)
}

best_k_index <- which.max(accuracies)
best_k <- k_values[best_k_index]

final_pred <- knn(train = train_scaled,
                  test = test_scaled,
                  cl = train_outcome,
                  k = best_k)
conf_matrix <- table(Predicted = final_pred, Actual = test_outcome)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
conf_matrix
paste("Best k:", best_k)
paste("Accuracy:", round(accuracy, 4))
```


```{r knn2}
cv_results <- data.frame(
  k = k_values,
  Accuracy = accuracies
)

ggplot(cv_results, aes(x = k, y = Accuracy)) +
  geom_line(color = "black", size = 1) +
  geom_point(color = "black", size = 2) +
  labs(title = "KNN Accuracy vs. k",
       subtitle = paste("Best k =", best_k),
       x = "k (# of Neighbors)",
       y = "Accuracy") 

```
By comparing the accuracy of the model for each k value we are able to find the ideal k value for the model. We will now perform 10 fold cross-validation to improve this model even further. Cross-validation takes more data into account for each iteration and provides more reliable end results. 
```{r knn3}
train_df_scaled <- as.data.frame(train_scaled)
train_df_scaled$Type <- as.factor(train_outcome)

ctrl <- trainControl(method = "cv", number = 10)

set.seed(77)
knn_cv <- train(
  Type ~ ., data = train_df_scaled,
  method = "knn",
  tuneGrid = data.frame(k = 1:20),
  trControl = ctrl
)
knn_cv$bestTune
knn_cv$results

ggplot(knn_cv) +
  labs(title = "10-Fold CV Accuracy for KNN", x = "k (Number of Neighbors)", y = "Accuracy")

```
After performing 10 fold cross-validation, we see a new ideal value for k. This is expected as cross-validation tests the data using multiple folds. Even though the accuracy for this model is lower, it should be the best as cross-validation evaluates performance on multiple train/test splits.

## Logistic Regression
We will now create a simple logistic regression model to compare to our tuned K-Nearest Neighbors model. Ideally, our KNN model will have a higher accuracy, showing that our efforts have paid off. 
```{r log}
log_model <- glm(Type ~ ., data = trainDF, family = binomial)

# Predict probabilities on test set
log_probs <- predict(log_model, newdata = testDF, type = "response")

# Convert probabilities to class labels using 0.5 threshold
log_pred <- ifelse(log_probs > 0.5, 1, 0)
log_pred <- factor(log_pred, levels = c(0, 1))

# Confusion matrix and accuracy
log_conf_matrix <- table(Predicted = log_pred, Actual = testDF$Type)
log_accuracy <- sum(diag(log_conf_matrix)) / sum(log_conf_matrix)

log_conf_matrix
paste("Logistic Regression Accuracy:", round(log_accuracy, 4))
```
## Conclusion
All in all, we achieved high accuracy for all 3 classification models. The first KNN model had the highest accuracy, the cross-validated KNN model had the second best, while logistic regression had the lowest. Both K-Nearest Neighbors having higher accuracy than logistic regression shows that this model is effective and we have tuned it well. Even though the first model has the highest accuracy, I must recommend using the KNN model with cross-validation. While the accuracy for this model is slightly lower, cross-validation evaluates performance across multiple more splits, reduces variance, and gives a more generalized model. As a result, the K-Nearest neighbors classification model with 10 fold cross-validation is my best recommendation for future endeavors. 















